---
output:
  html_document: default
  pdf_document: default
---
% DS 201 (Lab 1)
% Sidak Yntiso <sgy210@nyu.edu>, Office Hours: CDS 7th floor, 10-12 pm
% February 3, 2020



# Game plan
- This presentation, created using [knitr](http://yihui.name/knitr/), is available on the course website
- We will cover the basics (no familarity required): data structures, conditions + loops, data manipulation (dplyr, data.table)

# R Resources
- [The Art of R Programming](http://heather.cs.ucdavis.edu/~matloff/132/NSPpart.pdf) - N. Matloff
- [Modern Applied Statistics with S](http://www.stats.ox.ac.uk/pub/MASS4/) - W. Venables and B. Ripley
- [Advanced R Programming](http://adv-r.had.co.nz/) - H. Wickham
- [The R Inferno](http://www.burns-stat.com/documents/books/the-r-inferno/) - P. Burns
- Reading R documentation: <a href="http://stat.ethz.ch/R-manual/R-patched/library/stats/html/lm.html" target="_blank">`?lm`</a>
- CRAN documentation: <a href="http://cran.r-project.org/web/packages/AER/index.html" target="_blank">AER</a>
- Journal documentation: <a href="http://www.jstatsoft.org/v27/i02/">`plm`</a>
- Resources for R Markdown: (https://bookdown.org/yihui/rmarkdown/)

# Preliminaries
- You will need to install [R](https://repo.miserver.it.umich.edu/cran/) and an environment e.g. [Rstudio](https://rstudio.com/)
- Before running file, install knitr: install.packages("knitr")

```
require(knitr)
knit(Lab1.Rmd)
```

# Data Structures
```{r}
# 3 types of data structures: vectors, arrays, and data frames
#Two types of vectors: atomic/lists
#Difference: atomic vectors are flat and will contain only elements of same type
rm(list = ls())
atom <- c(nums = 1:5,c(1),lets=letters[1:5])
str(atom)

lst <- list(nums = 1:5,c(1),lets=letters[1:5])
str(lst)

#three ways to index lists
lst$nums
lst[[1]]
lst["nums"]

#combine vectors with c
c(lst,lst)

#convert lst to atomic
str(unlist(lst))

##Arrays e.g. matrices
mat <- matrix(rnorm(6),nrow=6,ncol=1) #rnorm, runif,rpois, etc generate random variables
mat[,1]
mat[1,]

str(mat)
ncol(mat)
nrow(mat)

#add/remove columns
mat <- cbind(1,mat)
rbind(mat,mat)

#matrices are easy to manipulate
#transpose
t(mat)
#X'X
(t(mat) %*% mat)
#(X'X)^-1
solve((t(mat) %*% mat))
#OLS
y <- 2 + 2*mat[,2] + rnorm(6)
solve(t(mat) %*% mat) %*% t(mat) %*% y
 
##Dataframes are matrices that allow mixing different types
dat <- data.frame(mat)
dat$lst <- lst
str(dat)

#rename columns or rows
colnames(dat) <- letters[1:ncol(dat)]
rownames(dat) <- c(1:nrow(dat))

```


# Conditions and Loops
```{r}


#loops are easy to write but can be slow. vectorize when possible
for (st in c(1:length(state.abb))){
  print(state.abb[st])
}


#functions
sum_US_area <- function(state.abb,state.area){
  US_area <- 0
  for (st in c(1:length(state.abb))){
    US_area <- US_area + state.area[st]
  }
  return(US_area)
}
sum_US_area(state.abb,state.area)

#useful functions you will likely use
print(
  c("print", "cat", "paste", "with", "length", "sort", "order", "unique", "rep", "nrow", "ncol",
    "complete.cases", "subset", "merge", "mean", "sum", "sd", "var", "lag","lm", "model.matrix","coef", "vcov",
    "residuals", "vcovHC (from sandwich)", "ivreg (from AER)", "summary", "pdf", "plot")
)


#use apply, lapply, sapply for similar reasons
#apply (aptly named) applies some function to rows (1) or columns (2) of matrix
head(state.x77)
apply(state.x77,2,mean)

#sapply applies some function to every element of a vector
sapply(mat, function(x) x+1)
#lapply is the same as sapply except it returns list output (instead of vector output)
class(sapply(mat, function(x) x+1))
class(lapply(mat, function(x) x+1))


#if/else
if ("NY" %in% state.abb){
  print("NY is a state")
} else {
  print("NY is not a state")
}

```

# Data Manipulation
```{r}
data(iris)
#to load datasets use haven::read_dta,read.table,data.table::fread, etc

#Many ways to handle data in R. base (above), dplyr and data.table (below)

#dplyr syntax
#install.packages("dplyr")
library(dplyr)
#summarize data by column
iris %>% 
  summarise(av_sepal_len=mean(Sepal.Length))
#summarize data by column and group
iris %>%  
  group_by(Species) %>% 
  summarise(av_sepal_len=mean(Sepal.Length))
#classify data by column and group
iris %>%  
  group_by(Species) %>% 
  mutate(high_sep_len=if_else(Sepal.Length > mean(Sepal.Length),"High","Low"))
#creating new data
iris_average <- iris %>%  
    group_by(Species) %>% 
    summarise(av_sepal_len=mean(Sepal.Length))

#data.table synatx 
#install.packages(data.table)
#data.table is faster with larger datasets
library(data.table)
iris <- data.table(iris)

#subsetting data by selecting rows
head(iris[Species == 'virginica'])

#selecting columns. the dot symbol calls lists
iris[, .(Species)]

#selecting rows and columns. all columns except Petal.Length. with=F 
head(iris[Species == 'virginica', -c("Petal.Length")])

#computations
iris[Species == 'virginica', .( mean_petal_length = mean(Petal.Length)) ]

#computations by group
iris[, .( mean_petal_length = mean(Petal.Length),
                                sd_petal_length = sd(Petal.Length),
                                N = .N),
     by = .(Species)]

```

