---
title: 'Lab 9: Difference in Difference Estimators'
author: "Sidak Yntiso sgy210@nyu.edu"
date: "April 06, 2020"
output:
  beamer_presentation:
  incremental: true
---
  
# Identifying assumptions
  
$Y_{it} = \pi D_{it}+ X'_{it} \beta_i + \alpha_i + \delta_t + \epsilon_{it}$

- Full rank regression matrix (variation in $X$)
-  Zero conditional mean of the errors: $\mathbb{E}[\epsilon_{it}|D_{it},X_{it},\alpha_i,\delta_t]=0$ for $t=1,..,T$

. . .

- Conditional independence of errors: $Cov(\epsilon_{it}, \epsilon_{jt} | D_{it},X_{it},\alpha_i,\delta_t) = 0$ 
- Homoskedasticity of the errors: $Var(\epsilon_{it} | D_{it},X_{it},\alpha_i,\delta_t) = \sigma^2$
  - or cluster robust standard errors or block bootstrap

# Empirical illustrations

Paper 1

- Multiple time periods; same treatment initiation period
- Visualizing parallel trends

. . .

Paper 2

- Multiple time periods; different treatment initiation periods
- Effect hetereogeneity; measurement error


# Paper 1

Effect of Medicaid expansion (Sommers et al 2012 (NEJM))

- What is the effect of expanded adult Medicaid eligibility 
- Expansion states (New York, Maine, and Arizona) passed new laws in 2000/01 
- Comparison group is neighboring states without expansions.
- Outcome is disease-related county-level mortality from the CDC

DiD Design

- first difference: expansion states and nonexpansion states 
- second difference: before and after reform


# First difference

```{r paper1-data}
rm(list=ls())
library(estimatr)
medicaid_study = haven::read_dta("medicaid_study.dta")

#what's the mean death rate for expansion 
#states in post-reform period (after 2001)
lm2005 <- lm_robust(cruderate~MedicaidExpansion,
                    subset(medicaid_study,year==2005))
summary(lm2005)$coefficients[,c(1:3)]
```


# Difference-in-means estimates by year
```{r paper1-fd}
dim_estimates <- c()
se_estimates <- c()
#what's the mean death rate by treatment
for (j in c(2002:2016)){
  lmj <- lm_robust(cruderate~MedicaidExpansion,
                   subset(medicaid_study,year==j))
  dim_estimates <- c(dim_estimates,
                     summary(lmj)$coefficients[2,1])
  se_estimates <- c(se_estimates,
                    summary(lmj)$coefficients[2,2])
}
#store results for years 2002-2016
dat1 <- data.frame(year=c(2002:2016),
                  dim_estimates = dim_estimates,
                  se_estimates = se_estimates)
```

# Difference-in-means estimates by year plot
```{r paper1-fd-plot}
interval2 <- -qnorm((1-0.95)/2)  # 95% multiplier
#plot the effects by year
library(ggplot2)
p <- ggplot(aes(x=year,y=dim_estimates),data=dat1)+
  geom_hline(yintercept = 0, colour = gray(1/2), lty = 2)+
  geom_point(aes(x = year, y = dim_estimates),lwd = 2)+
  geom_linerange(aes(x = year, 
                     ymin = dim_estimates - 
                       se_estimates*interval2,
                     ymax = dim_estimates + 
                       se_estimates*interval2),
                 lwd = 1)+xlab("Year")+theme_bw()+
  ylab("Effect of Expansion")
```

# Visualize
```{r paper1-first_diff_plot, fig.cap='', fig.height=4, fig.width=6, warning=FALSE}
p
```


# Second difference
```{r paper1-second_diff}
#identify treatment states
medicaid_study$trt_state = 0
medicaid_study$trt_state[medicaid_study$state%in% 
    c("New York","Maine","Arizona")] = 1

#what's the mean cruderate in post-reform period
lm2001 <- lm_robust(cruderate~I(year>=2001),medicaid_study)
summary(lm2001)$coefficients[,c(1:3)]
```

# Second difference plot
```{r paper1-pl-trends, message=FALSE, warning=FALSE}
#plot trends in death rate by treatment group
p2 <- ggplot(medicaid_study, aes(year, cruderate, 
                color = factor(trt_state)) )  +
  stat_summary(aes(group = factor(trt_state)),
               geom="line") +
  geom_vline(xintercept = 2001, colour = gray(1/2))+
  theme_bw()
```


# Visualizing Parallel Trends


```{r paper1-second_diff_plot, fig.cap='', fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
p2
```

# Difference in differences
```{r paper1-did}
#two way fixed effects estimate
m1 <- lm_robust(cruderate ~ MedicaidExpansion+
                   as.factor(year)+as.factor(countycode), 
                 clusters=countycode,data = medicaid_study)
summary(m1)$coefficients[c(1:2),c(1:3)]
```

# Testing Parallel Trends Setup
```{r paper1-pll_test}
#generate placebo treatments
for (j in c(1999:2001,2003:2016)){
  assign(paste("treat",j,sep=""),
       medicaid_study$trt_state*
         I(medicaid_study$year==j))
}
#DiD model with placebo treatments
m2 <- formula(paste("cruderate~",
  paste("treat",c(1999:2001,2003:2016),
        sep="",collapse="+" ),
  "+as.factor(year)+as.factor(countycode)",sep=""))
#running the model
m2 <- lm_robust(m2,data = medicaid_study, 
                 clusters=countycode)
```

# Placebo DiD estimates data
```{r message=FALSE, warning=FALSE}
#storing the placebo DiD estimates
dim_estimates <- c(); se_estimates <- c()
for (j in c(2:18) ){
  dim_estimates <- c(dim_estimates,
                     summary(m2)$coefficients[j,1])
  se_estimates <- c(se_estimates,
                    summary(m2)$coefficients[j,2])
}
#saving to a dataset
dat2 <- data.frame(year=c(1999:2001,2003:2016),
                  dim_estimates = dim_estimates,
                  se_estimates = se_estimates)
```

# Placebo DiD estimates plot
```{r paper1-pll-test-setup}
#plotting the placebo DiD estimates by year
p3 <- ggplot(aes(x=year,y=dim_estimates),data=dat2)+
  geom_hline(yintercept = 0, colour = gray(1/2), lty = 2)+
  geom_point(aes(x = 2002, y = 0),lwd = 2,colour="red")+
  geom_point(aes(x = year, y = dim_estimates),lwd = 2)+
  geom_linerange(aes(x = year, 
                     ymin = dim_estimates - 
                       se_estimates*interval2,
                     ymax = dim_estimates + 
                       se_estimates*interval2),
                 lwd = 1)+xlab("Year")+theme_bw()+
  ylab("Placebo DiD Effects")
```


# Visualize
```{r paper1-pll-test-plot, fig.cap='', fig.height=4, fig.width=6, warning=FALSE}
p3
```

# Paper 2

Voter Identification laws: require government ID to vote (Hajnal et al (2017) (HLN) and Grimmer et al 2018)

- Strict ID states are Arizona (2004), Georgia (2005), Indiana (2005), Kansas (2011), Mississippi (2011), North Dakota (2013), Ohio (2006), Tennessee (2011), Texas (2011), Virginia (2012), and Wisconsin (2016)
- Minority voters: much less likely to hold IDs (Ansolabehere and Hersh 2016)
- What is effect of ID laws on turnout? by race?

. . .

DiD Design

- Cooperative Congressional Election Study (2006-2014)
- Dependent Variable: General/Primary Election Turnout
- Treatment: Strict Voter ID Law in state
- first difference: states with strict voter rules versus others 
- second difference: year before and after reform


# Data

```{r paper2-data}
##Loading the data from HLN 
dd<- read.delim('HKL_V2_data.tab', sep='\t')
#list of covariates to condition on
covs <- c("foreignb","firstgen","age","educ","inc",
         "male","married","childrenz","unionz",
         "unemp","ownhome","protestant","catholic",
         "jewish","atheist","days_before_election",
         "early_in_person","vote_by_mail",
         "no_excuse_absence_",
         "presidentialelectionyear",
         "gubernatorialelectionyear",
         "senateelectionyear",
         "marginpnew","newstrict")
```

# Model

```{r message=FALSE, warning=FALSE}
#model
m1 <- formula(paste("votegenval~stricty+black+hispanic +
              asian + mixedrace + blackstricty + 
              hispstricty + asianstricty +
              mixedracestricty + as.factor(year) + 
              as.factor(state)+",paste(covs,collapse="+")))
###Replication Column 1, Table A9 HLN (no clustering)
model1<- lm_robust(m1,data = subset(dd,voteregpre==1))
```

# Estimates
```{r paper2-estimates, message=FALSE, warning=FALSE}
#store main and interaction results
results <- data.frame(
  race = c("White","Black","Hispanic","Asian","Mixed"))
 results$m1_ses <- results$m1_est <-NA
results$m1_est[1] = 
  summary(model1)$coefficients['stricty',1]
results$m1_est[2] = results$m1_est[1] +
  summary(model1)$coefficients['blackstricty',1]
results$m1_est[3] = results$m1_est[1] +
  summary(model1)$coefficients['hispstricty',1]
results$m1_est[4] = results$m1_est[1] +
  summary(model1)$coefficients['asianstricty',1]
results$m1_est[5] = results$m1_est[1] +
  summary(model1)$coefficients['mixedracestricty',1]
```

# Standard errors 
```{r paper2-ses}
#variance covariance matrix of coefficients
v_model<- vcov(model1) 
#variance of main+interaction effects 
#using Var(X+Y) = Var(X)+ Var(Y)+ 2*Cov(XY)
results$m1_ses[1]<- sqrt(v_model['stricty', 'stricty'])
results$m1_ses[2]<- sqrt(v_model['stricty', 'stricty'] 
                  +v_model['blackstricty', 'blackstricty'] 
                  +2*v_model['stricty', 'blackstricty'])
results$m1_ses[3]<- sqrt(v_model['stricty', 'stricty'] 
                  +v_model['hispstricty', 'hispstricty'] 
                  +2*v_model['stricty', 'hispstricty'])
results$m1_ses[4]<- sqrt(v_model['stricty', 'stricty'] 
                  +v_model['asianstricty', 'asianstricty'] 
                  +2*v_model['stricty', 'asianstricty'])
results$m1_ses[5]<- sqrt(v_model['stricty', 'stricty'] 
                  +v_model['mixedracestricty','mixedracestricty']
                  +2*v_model['stricty', 'mixedracestricty'])
```

# Summarize results
```{r paper2-results}
#t statistic
results$tval <- results$m1_est/results$m1_ses
results
```

# What went wrong

How could it be that "Racial and ethnic minorities . . . are especially hurt by strict voter identification laws" (HLN) if voter id laws do not suppress turnout?

. . .

Although they control for a certain type of omitted variable, fixed-effects estimates are notoriously susceptible to measurement error. 

- On one hand, outcomes tend to be persistent (e.g. union membership).
- On the other hand, measurement error often changes from year-to-year (e.g. union status may be misreported
or miscoded this year but not next year).
- while union status may be misreported or miscoded for only a few workers in any single year, the observed year-to-year changes in union status may be mostly noise. 

# Virginia
State policy in Virginia (in effect till 2010)

- No access to vote history
- HLN code VA CCES respondents as nonvoters in 2006,2008.
- In fact their vote status are missing

```{r paper2-va, fig.show='hide'}
out<- which(dd$state=='Virginia') #VA data
#VA turnout by year
vas<- table(dd$votegenval[out], dd$year[out])
vas
```


# Dropping VA
```{r paper2-droppingva}
##identifying states
un_state<- unique(dd$state)
un_state<- sort(un_state)
states<- as.character(rev(un_state))
##storing the models
state_early<- list()
sample_states = sample(states,size = 10)
if (!"Virginia" %in% sample_states){
  sample_states[11] = "Virginia"
}
for(z in sample_states){
  ##dropping one state at a time
  drops<- rep(0, nrow(dd))
  drops[which(dd$state==z)]<- 1
  temp<- lm_robust(m1,data =subset(dd,
          voteregpre==1&drops==0))
  state_early[[z]]<- temp
}
```


# Store results
```{r paper2-droppingva-estimates, message=FALSE, warning=FALSE}
#store main and interaction results
White<- Black<- Hispanic<- Asian<- c()

for(z in 1:length(sample_states)){
White[z]<-summary(state_early[[z]])$
  coefficients['stricty', 1]

Black[z]<- White[z]+summary(state_early[[z]])$
  coefficients['blackstricty',1]

Hispanic[z]<- White[z]+summary(state_early[[z]])$
  coefficients['hispstricty',1]

Asian[z]<- White[z]+summary(state_early[[z]])$
  coefficients['asianstricty',1]
}

```

# Results
```{r paper2-droppingvaresults, fig.height=3, fig.width=5}
boxplot(cbind(White, Black, Hispanic, Asian), cex = 2,
        main = "Boxplot of Effect Estimates Shows 
        \nVirginia Exerts Substantial Influence", pch=20)
abline(h = 0, lwd = 3, lty = 2, col=gray(0.5))
```

# Final Estimates
```{r paper2-corrected-model}
###Finally, dropping Virginia and 2006 observations
good_obs<- rep(1, nrow(dd))
good_obs[which(dd$year==2006)]<- 0
good_obs[which(dd$year==2008 & dd$state=='Virginia')]<- 0 


###running the regression
model2<- lm_robust(m1,data = 
                     subset(dd,voteregpre==1& good_obs==1))
```

# Storing results
```{r paper2-corrected-estimates, message=FALSE, warning=FALSE}
#store main and interaction results
results <- data.frame(
  race = c("White","Black","Hispanic","Asian","Mixed"))
results$m2_ses <- results$m2_est <-  NA
results$m2_est[1]<-
  summary(model2)$coefficients['stricty', 1]

results$m2_est[2]<-results$m2_est[1]+
  summary(model2)$coefficients['blackstricty', 1]

results$m2_est[3]<-results$m2_est[1]+
  summary(model2)$coefficients['hispstricty', 1]

results$m2_est[4]<-results$m2_est[1] +
  summary(model2)$coefficients['asianstricty', 1]

results$m2_est[5]<-results$m2_est[1] +
  summary(model2)$coefficients['mixedracestricty', 1]
```

# Standard Errors
```{r paper2-corrected-ses}
#variance covariance matrix of coefficients
v_model<- vcov(model2) 
#variance of main+interaction effects 
results$m2_ses[1]<- sqrt(v_model['stricty', 'stricty'])
results$m2_ses[2]<- sqrt(v_model['stricty', 'stricty']
                  +v_model['blackstricty', 'blackstricty'] 
                  +2*v_model['stricty', 'blackstricty'])
results$m2_ses[3]<- sqrt(v_model['stricty', 'stricty'] 
                  +v_model['hispstricty', 'hispstricty'] 
                  +2*v_model['stricty', 'hispstricty'])
results$m2_ses[4]<- sqrt(v_model['stricty', 'stricty'] 
                  +v_model['asianstricty', 'asianstricty'] 
                  +2*v_model['stricty', 'asianstricty'])
results$m2_ses[5]<- sqrt(v_model['stricty', 'stricty'] 
                   +v_model['mixedracestricty','mixedracestricty']
                   +2*v_model['stricty', 'mixedracestricty'])
```

# Results
```{r paper2-correctedresults}
#t statistic
results$tval <- results$m2_est/results$m2_ses
results
```

# Block bootstrap

- *Block bootstrap* is when we bootstrap whole groups (states, etc) instead of $it$ pairs. 
- Accounts for correlations within the groups (serial correlation, etc).
- Computationally tricky because you need to keep track of all the multilevel indices. 

# Block bootstrap coding
```{r block-boostrap, message=FALSE, warning=FALSE}
library(dplyr)
#simplify data for bootstrap
dat <- subset(dd,voteregpre==1)  %>% 
  group_by(year,state) %>% 
  summarize(votegenval = mean(votegenval,na.rm = T),
            stricty = max(stricty))
#running the model
model4 <- lm_robust(votegenval ~ stricty + as.factor(year) +
              as.factor(state),data =dat,clusters=as.factor(state))

#Trick to get the indices for each group
lookup <- split(1:nrow(dat), dat$state)
names(lookup[10]); head(lookup[[10]]) #Inspect 10th state
gnames <- names(lookup) #extract state names
```

# Block bootstrap
```{r block-boostrap-code, message=FALSE, warning=FALSE}
sims = 2000
ates <- c()
for (j in 1:sims){
  #sample from states
  star <- sample(gnames, size = length(gnames),
                 replace = TRUE)
  dat.star <- dat[unlist(lookup[star]),] #new data
  temp<- lm_robust(votegenval ~ stricty + as.factor(year) +
              as.factor(state),data =dat.star)
  ates[j] <- summary(temp)$coefficients['stricty',1]
}
sd(ates)
summary(model4)$coefficients['stricty',2]
```
